---
title: "Exercise 12"
author: "Jennie"
date: "2024-04-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

# Exercise 12

## **Learning Objectives**

-   Run exploratory multivariate regression to evaluate different models

#### **Preliminaries**

-   Set up a new ***GitHub*** repo in your ***GitHub*** workspace named “exercise-12” and clone that down to your computer as a new ***RStudio*** project. The instructions outlined as **Method 1** in [**Module 6**](https://difiore.github.io/ada-2024/06-module.html) will be helpful.

Using the {tidyverse} `read_tsv()` function, load the “Mammal_lifehistories_v2.txt” dataset from [this URL](https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Mammal_lifehistories_v2.txt) as a “tibble” named **d**. As discussed in class, this is dataset that compiles life history and other variables for over 1400 species of placental mammals from 17 different Orders. **Data source**: Ernest SKM. (2003). Life history characteristics of placental nonvolant mammals. *Ecology* 84: 3402–3402.

```{r Preliminaries.1}
# load dataset from URL
library(tidyverse)
f <- "https://raw.githubusercontent.com/difiore/ada-2024-datasets/main/Mammal_lifehistories_v2.txt"
d <- read_tsv(f, col_names = TRUE)
```

-   Do a bit of exploratory data analysis with this dataset, e.g., using the {skimr} package. Which of the variables are categorical and which are numeric?

The categorical variables are **order**, **family**, **Genus**, and **species**. The numeric variables are **mass(g)**, **gestation(mo)**, **newborn(g)**, **weaning(mo)**, **wean mass(g)**, **AFR(mo)**, **max. life(mo)**, **litter size**, **litters/year**, and **refs**.

```{r Preliminaries.2}
# view a summary of the columns (types, min, max, mean, etc.)
library(skimr)
skim(d)
```

## **Challenge**

#### **Step 1**

-   Replace all values of -999 (the authors’ code for missing data) with `NA`.

> **HINT:** This is easy to do in base {R}, but you can also check out the `replace_with_na_all()` function from the {naniar} package.

```{r Step.1}
# this dataset uses "-999.0" to mean "NA"; make it say "NA" instead
library(naniar)
d_new <- replace_with_na_all(data = d, condition = ~.x == -999.0)
```

#### **Step 2**

-   Drop the variables **litter size** and **refs**.

```{r Step.2}
# litter size = 12th column; refs = 14th column
d_new <- d_new[ -c(12, 14)]
```

#### **Step 3**

-   Log transform **all** of the other numeric variables.

> **HINT:** There are lots of ways to do this… look into `mutate(across(where(), .funs))` for an efficient motif.

```{r Step.3}
# log transform the numeric variables
d_new <- d_new |> 
  mutate(across(where(is.numeric), log))
# this gives warnings that NaNs were produced, but the log values appear to be okay.
```

#### **Step 4**

-   Regress the (now log transformed) age [**gestation(mo)**, **weaning(mo)**, **AFR(mo)** (age at first reproduction), and **max. life(mo)** (maximum lifespan)] and mass [**newborn(g)** and **wean mass(g)**] variables on (now log transformed) overall body **mass(g)** and add the residuals to the dataframe as new variables (**relGest**, **relWean**, **relAFR**, **relLife**, **relNewbornMass**, and **relWeaningMass**).

> **HINT:** Use “na.action=na.exclude” in your`lm()` calls. With this argument set, the residuals will be padded to the correct length by inserting NAs for cases with missing data.

```{r Step.4}
# I first tried to run one of the models below, but it was giving an error that some of the values are NaN or Inf. Will replace those with NA:
d_new[is.na(d_new) | d_new=="Inf"] = NA
d_new[is.na(d_new) | d_new=="-Inf"] = NA
# And now residuals for the 6 listed variables:
relGest <- resid(lm(`gestation(mo)` ~ `mass(g)`, data = d_new, na.action = na.exclude))
relWean <- resid(lm(`weaning(mo)` ~ `mass(g)`, data = d_new, na.action = na.exclude))
relAFR <- resid(lm(`AFR(mo)` ~ `mass(g)`, data = d_new, na.action = na.exclude))
relLife <- resid(lm(`max. life(mo)` ~ `mass(g)`, data = d_new, na.action = na.exclude))
relNewbornMass <- resid(lm(`newborn(g)` ~ `mass(g)`, data = d_new, na.action = na.exclude))
relWeaningMass <- resid(lm(`wean mass(g)` ~ `mass(g)`, data = d_new, na.action = na.exclude))
# Add them to d_new:
d_new$relGest <- relGest
d_new$relWean <- relWean
d_new$relAFR <- relAFR
d_new$relLife <- relLife
d_new$relNewbornMass <- relNewbornMass
d_new$relWeaningMass <- relWeaningMass
```

#### **Step 5**

-   Plot residuals of max lifespan (**relLife**) in relation to **Order**. Which mammalian orders have the highest residual lifespan?

Primates, Scandentia, and Proboscidea appear to have the highest **relLife**. Using the `summary` function, the mean relLife for Primates is 0.7379 and Proboscidea is 0.1595, but Scandentia is mysteriously missing from the summary.

-   Plot residuals of newborn mass (**relNewbornMass**) in relation to **Order**. Which mammalian orders have the have highest residual newborn mass?

Macroscelidea, Scandentia, and Primates appear to have the highest **relNewbornMass**.

-   Plot residuals of weaning mass (**relWeaningMass**) in relation to **Order**. Which mammalian orders have the have highest relative weaning mass?

Insectivora and Perissodactyl appear to have the highest **relWeaningMass**.

> **NOTE:** There will be lots of missing data for the latter two variables!

```{r Step.5}
library(ggplot2)
# relLife in relation to Order:
# plot:
plot1 <- ggplot(data = d_new, aes(x = `order`, y = relLife)) + geom_boxplot() + geom_jitter(alpha = 0.1) + theme(axis.text.x=element_text(angle = -90, hjust = 0))
print(plot1)
# summary:
tapply(d_new$relLife, d_new$order, summary)

# relNewbornMass in relation to Order:
# plot:
plot2 <- ggplot(data = d_new, aes(x = `order`, y = relNewbornMass)) + geom_boxplot() + geom_jitter(alpha = 0.1) + theme(axis.text.x=element_text(angle = -90, hjust = 0))
print(plot2)
# summary:
tapply(d_new$relNewbornMass, d_new$order, summary)

# relWeaningMass in relation to Order:
# plot:
plot3 <- ggplot(data = d_new, aes(x = `order`, y = relWeaningMass)) + geom_boxplot() + geom_jitter(alpha = 0.1) + theme(axis.text.x=element_text(angle = -90, hjust = 0))
print(plot3)
# summary:
tapply(d_new$relWeaningMass, d_new$order, summary)
```

#### **Step 6**

-   Run models and a model selection process to evaluate what (now log transformed) variables best predict each of the two response variables, **max. life(mo)** and **AFR(mo)**, from the set of the following predictors: **gestation(mo)**, **newborn(g)**, **weaning(mo)**, **wean mass(g)**, **litters/year**, and overall body **mass(g)**.

> **HINT:** Before running models, winnow your dataset to drop rows that are missing the respective response variable or any of the predictors, e.g., by using `drop_na()`.

-   For each of the two response variables, indicate what is the best model overall based on AICc and how many models have a delta AICc of 4 or less.

For the model where the response variable is **max. life(mo)**, there are 9 models with a delta AICc of 4 or less. The best one includes two predictors, **gestation(mo)** and **weaning(mo)**. For the model where the response variable is **AFR(mo)**, there are 17 models with a delta AICc of 4 or less, with the best model including two predictors, **litters/year** and **weaning(mo)**.

-   What variables, if any, appear in all of this set of “top” models?

In the model with **max. life(mo)** as the response variable, **weaning(mo)** appears in all 9 of the best models. In the model where **AFR(mo)** is the response variable, no variable appears in all 17 of the best models, but one comes close: **weaning(mo)** appears in 15 of 17.

-   Calculate and plot the model-averaged coefficients and their CIs across this set of top models.

```{r Step.6}
# Drop rows with missing data:
d_new <- d_new |> drop_na(`max. life(mo)`, `AFR(mo)`, `gestation(mo)`, `newborn(g)`, `weaning(mo)`, `wean mass(g)`, `litters/year`, `mass(g)`)

# Model for max. life(mo):
m_Life <- lm(data = d_new, `max. life(mo)` ~ `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, na.action = "na.fail")
# Model selection:
library(MuMIn)
mods_Life <- dredge(m_Life, beta = "none")
mods.avg_Life <- summary(model.avg(mods_Life, subset = delta < 4, fit = TRUE))
print(mods.avg_Life)
# Confidence intervals:
confint_Life <- confint(mods.avg_Life)
print(confint_Life)
plot4 <- plot(mods.avg_Life, full = TRUE, intercept = FALSE)
print(plot4)

# Model for AFR(mo):
m_AFR <- lm(data = d_new, `AFR(mo)` ~ `gestation(mo)` + `newborn(g)` + `weaning(mo)` + `wean mass(g)` + `litters/year` + `mass(g)`, na.action = "na.fail")
# Model selection:
mods_AFR <- dredge(m_AFR, beta = "none")
mods.avg_AFR <- summary(model.avg(mods_AFR, subset = delta < 4, fit = TRUE))
print(mods.avg_AFR)
# Confidence intervals:
confint_AFR <- confint(mods.avg_AFR)
print(confint_AFR)
plot5 <- plot(mods.avg_AFR, full = TRUE, intercept = FALSE)
print(plot5)
```
